Dependent Types for Safe and Secure Web Programming

### 2014-01-29 paper/effects.tex
(Hopefully) final revision of paper before IFL2013 submission.
Fix error introduced in last commit

\noindent
\texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
which describes the context in which the effectful program will be run, a
list of side effects \texttt{es} that the program is permitted to use, and the
programs return type \texttt{a}. The name \texttt{m} for the computation context is
suggestive of a monad, but there is no requirement for it to be so.

@@

Effects such as state and exception are described as algebraic data types,
and run by giving \emph{handlers} for specific computation contexts. 
Effects have a corresponding \emph{resource} (in the case of state, the
resource is simply the current state). Executing an effectful operation may
change the resource and return a value:

@@

Figure \ref{fig:fileeffect} shows how the protocol is encoded as an
effect.
Note that the types of the input and output resources describes how resource
state changes in each operation: opening a file may fail, so
changes an empty resource to
a resource containing either a unit or an open file; 
reading a line is only possible if the
resoource is a file open for reading, etc.
The handler for this effect for an \texttt{IO} computation context will
execute the required primitive I/O actions.

@@

\noindent
The type of \texttt{testFile} states
that File I/O and console I/O are available effects, and in particular that
the resource associated with the File I/O will be in the same state on entry
and exit. We use \texttt{if\_valid} to handle possible failure---this is
a function provided by the \texttt{Effects} library which checks whether
a resource of type \texttt{Either a b} indicates failure (\texttt{Left a})
or success (\texttt{Right b}) and branches and updates the resource
accordingly.
Therefore, attempting to write to the file, failing to check for
success, or failing to open or close the
file, would cause a \emph{compile-time} error.

We will use this technique extensively
throughout this paper: describe a resource usage protocol in terms of
state transitions; implement an effect which captures that protocol; implement
programs which, by using this effect, implicitly carry a proof that the resource
protocol has been correctly followed.

>>>>

\noindent
\texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
which describes the context in which the effectful program will be run, a
list of side effects \texttt{es} that the program is permitted to use, and the
program's return type \texttt{a}. The name \texttt{m} for the computation context is
suggestive of a monad, but there is no requirement for it to be so.

@@

Effects such as `State' and `Exception' are described as algebraic data types,
and run by giving \emph{handlers} for specific computation contexts. 
Effects have a corresponding \emph{resource} (in the case of state, the
resource is simply the current state). Executing an effectful operation may
change the resource and return a value:

@@

Figure \ref{fig:fileeffect} shows how the protocol is encoded as an
effect.
Note that the types of the input and output resources describes how resource
state changes in each operation: opening a file may fail, so
changes an empty resource to
a resource containing either a unit or an open file; 
reading a line is only possible if the
resource is a file open for reading, etc.
The handler for this effect for an \texttt{IO} computation context will
execute the required primitive I/O actions.

@@

\noindent
The type of \texttt{testFile} states
that File I/O and console I/O are available effects, and in particular that
the resource associated with the File I/O will be in the same state on entry
and exit. We use \texttt{if\_valid} to handle possible failure---this is
a function provided by the \texttt{Effects} library which checks whether
a resource of type \texttt{Either a b} indicates failure (\texttt{Left a})
or success (\texttt{Right b}). In this case, the initial resource in the valid
branch is \texttt{OpenFile Read}, indicating that the file has been successfully 
opened for reading, whereas in the failing branch, the resource remains as \texttt{()}.

Attempting to write to the file, failing to check for
success, or failing to open or close the
file, would cause a \emph{compile-time} error.

We will use this technique extensively
throughout this paper: describe a resource usage protocol in terms of
state transitions; implement an effect which captures that protocol; and implement
programs which, by using this effect, implicitly carry a proof that the resource
protocol has been correctly followed.

### 2014-01-29 paper/introduction.tex
(Hopefully) final revision of paper before IFL2013 submission.
Fix error introduced in last commit

Many web applications are written in dynamically-checked scripting languages
such as PHP, Ruby or Python, which facilitate rapid development
\cite{w3techs:webpls}. However, such languages
do not provide the same static guarantees about runtime behaviour afforded by
programs with more expressive, static type systems, instead relying on
extensive unit testing to ensure correctness and security.

@@

\idris{} \cite{brady2013idris} is a language with full dependent types, and
extensive support for EDSLs through overloading and syntax macros. Through the
use of \idris{}, and a framework for describing resource protocols using
\emph{algebraic effects}~\cite{brady:effects}, we
present a dependently-typed web framework, which allows the construction of
programs with additional guarantees about correctness and security, whilst
minimising the increase in development complexity.

>>>>

Many web applications are written in dynamically-checked scripting languages
such as PHP, Ruby or Python, which facilitate rapid development
\cite{w3techs:webpls}. Such languages do not, however, provide the same static 
guarantees about runtime behaviour afforded by
programs with more expressive, static type systems, instead relying on
extensive unit testing to ensure correctness and security.

@@

\idris{} \cite{brady2013idris} is a language with full dependent types, and
extensive support for EDSLs through overloading and syntax macros. Through the
use of \idris{}, and a framework for describing resource protocols using
\emph{algebraic effects}~\cite{brady:effects}, we
present a dependently-typed web framework allowing the construction of
programs with additional guarantees about correctness and security, whilst
minimising the increase in development complexity.

### 2014-01-29 paper/main.tex
(Hopefully) final revision of paper before IFL2013 submission.
Fix error introduced in last commit

\begin{abstract}
Dependently-typed languages allow precise types to be used during development,
facilitating easier reasoning about the operation of programs. However,
stronger types bring a disadvantage that it becomes increasingly difficult to
write programs that are accepted by a type checker and additional proofs may
have to be specified by a programmer.

Embedded domain-specific languages (EDSLs) address this problem by introducing
a layer of abstraction over more specific underlying types, allowing
domain-specific code to be written in a high-level languages which uses
dependent types to enforce certain invariants without additional proof
obligations.

@@

%\keywords
%keyword1, keyword2

@@

\acks
This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC. The authors would like to thank contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
% SICSA / EPSRC (grant number?
% #idris
% Idris contributors, in particular Melissa for the SQLite bindings and whoever wrote Network.Cgi

>>>>

\begin{abstract}
Dependently-typed languages allow precise types to be used during development,
facilitating static reasoning about program behaviour. However,
stronger types bring a disadvantage that it becomes increasingly difficult to
write programs that are accepted by a type checker, meaning additional proofs may
have to be specified by a developer.

Embedded domain-specific languages (EDSLs) can help address this problem by
introducing a layer of abstraction over more precise underlying types,
allowing domain-specific code to be written in a high-level language which uses
dependent types to enforce invariants without imposing additional proof
obligations on an application developer.

@@

\keywords
Dependent types, web applications, security, verification
%keyword1, keyword2

@@

\acks
This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC. We would like to thank the contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
We are very grateful to Peter Thiemann and the anonymous reviewers for their insightful and constructive comments.
% SICSA / EPSRC (grant number?
% #idris
% Idris contributors, in particular Melissa for the SQLite bindings and whoever wrote Network.Cgi

### 2014-01-29 paper/protocols.tex
Fix error introduced in last commit

A call to \texttt{prepareStatement} attempts to create a prepared statement,
and a subsequent call to \texttt{beginExecution} allows data to be retrieved
from the database.
%
\texttt{CollectResults} operates on the row currently held in the buffer.
Firstly, the function checks
that there is a valid row in the buffer, and if so uses the
\texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data
from the database. This function is then called recursively until there are no
more rows to process.

@@

collectResults :
  EffM IO [SQLITE (Either (SQLiteExecuting InvalidRow) 
                          (SQLiteExecuting ValidRow))] 
          [SQLITE (SQLiteExecuting InvalidRow)] 
          (List (String, Int))
\end{SaveVerbatim}

\begin{SaveVerbatim}{collectres}
collectResults = 
  if_valid then do name <- getColumnText 1
                   age <- getColumnInt 2
                   step_res <- nextRow
                   xs <- collectResults
                   return $ (name, age) :: xs
  else return []

>>>>

A call to \texttt{prepareStatement} attempts to create a prepared statement,
and a subsequent call to \texttt{beginExecution} allows data to be retrieved
from the database.
%
\texttt{CollectResults} operates on the row currently held in the buffer.
Firstly, the function checks
that there is a valid row in the buffer, and if so uses the
\texttt{getColumnText} function to retrieve the data
from the database. This function is then called recursively until there are no
more rows to process.

@@

collectResults :
  EffM IO [SQLITE (Either (SQLiteExecuting InvalidRow) 
                          (SQLiteExecuting ValidRow))] 
          [SQLITE (SQLiteExecuting InvalidRow)] 
          (List (String, String))
\end{SaveVerbatim}

\begin{SaveVerbatim}{collectres}
collectResults = 
  if_valid then do name <- getColumnText 1
                   addr <- getColumnText 2
                   step_res <- nextRow
                   xs <- collectResults
                   return $ (name, addr) :: xs
  else return []
